---
title: "Capstone Project"
author: "Ryan Ramirez"
date: "2023-01-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview


Cyclistic is a substantial bike-share program that has multiple bike stations 
around the greater Chicago area. As the number of users grow, the company wants
to explore how they can increase the number of annual memberships. Specifically, 
we will focus on the behaviors of casual members to create recommendations that 
will encourage their transition into an annual member.


```{r install packages, include = FALSE}

library(tidyverse)
library("geosphere")
library("gridExtra") 
library("ggmap")
```

## Processing and Cleaning the Data


The data was gathered from a public source and made available by Motivate 
International Inc. found on the AWS cloud platform. 

Since this is public data,there is no personal info containing the riders bank 
details or home address.Each ride has a unique ride ID, however, the type of 
membership is shown to continue our analysis.


Prior to importing the data sets the following steps were done on Excel:

1. Calculated the ride duration for each ride in "hh:mm:ss" format
2. Added a column indicating the day of the week in numerical format



### Using R for data cleaning


The following was done with the imported csv files from Excel:

* The *ride_length* column was converted to a numeric double indicating seconds
* All NA's were omitted from each monthly dataset

The data includes rows where the station name or end staion names are missing,
but this will be analyzed later. 


```{r loading data, include = FALSE}
library(readr)
trip_2022_01 <- read_csv("C:/Users/kashi/OneDrive/Desktop/cyclistic/202201-divvy-tripdata/202201-divvy-tripdata.csv")
trip_2022_02 <- read_csv("C:/Users/kashi/OneDrive/Desktop/cyclistic/202202-divvy-tripdata/202202-divvy-tripdata.csv")
trip_2022_03 <- read_csv("C:/Users/kashi/OneDrive/Desktop/cyclistic/202203-divvy-tripdata/202203-divvy-tripdata.csv")
trip_2022_04 <- read_csv("C:/Users/kashi/OneDrive/Desktop/cyclistic/202204-divvy-tripdata/202204-divvy-tripdata.csv")
trip_2022_05 <- read_csv("C:/Users/kashi/OneDrive/Desktop/cyclistic/202205-divvy-tripdata/202205-divvy-tripdata.csv")
trip_2022_06 <- read_csv("C:/Users/kashi/OneDrive/Desktop/cyclistic/202206-divvy-tripdata/202206-divvy-tripdata.csv")
trip_2022_07 <- read_csv("C:/Users/kashi/OneDrive/Desktop/cyclistic/202207-divvy-tripdata/202207-divvy-tripdata.csv")
trip_2022_08 <- read_csv("C:/Users/kashi/OneDrive/Desktop/cyclistic/202208-divvy-tripdata/202208-divvy-tripdata.csv")
trip_2022_09 <- read_csv("C:/Users/kashi/OneDrive/Desktop/cyclistic/202209-divvy-tripdata/202209-divvy-publictripdata.csv")
trip_2022_10 <- read_csv("C:/Users/kashi/OneDrive/Desktop/cyclistic/202210-divvy-tripdata/202210-divvy-tripdata.csv")
trip_2022_11 <- read_csv("C:/Users/kashi/OneDrive/Desktop/cyclistic/202211-divvy-tripdata/202211-divvy-tripdata.csv")
trip_2022_12 <- read_csv("C:/Users/kashi/OneDrive/Desktop/cyclistic/202212-divvy-tripdata/202212-divvy-tripdata.csv")
```


```{r data cleaning, include = FALSE}
library(lubridate)
jan <- trip_2022_01
jan <- jan %>%
  mutate(ride_length, ride_length = as.numeric(ride_length))%>%
  na.omit(ride_length)

feb <- trip_2022_02
feb <- feb %>%
  mutate(ride_length, ride_length = as.numeric(ride_length))%>%
  na.omit(ride_length)

mar <- trip_2022_03
mar <- mar %>%
  mutate(ride_length, ride_length = as.numeric(ride_length))%>%
  na.omit(ride_length)

apr <- trip_2022_04
apr <- apr %>%
  mutate(ride_length, ride_length = as.numeric(ride_length))%>%
  na.omit(ride_length)

may <- trip_2022_05
may <- may %>%
  mutate(ride_length, ride_length = as.numeric(ride_length))%>%
  na.omit(ride_length)

jun <- trip_2022_06
jun <- jun %>%
  mutate(ride_length, ride_length = as.numeric(ride_length))%>%
  na.omit(ride_length)

jul <- trip_2022_07
jul <- jul %>%
  mutate(ride_length, ride_length = as.numeric(ride_length))%>%
  na.omit(ride_length)

aug <- trip_2022_08
aug <- aug %>%
  mutate(ride_length, ride_length = as.numeric(ride_length))%>%
  na.omit(ride_length)

sep <- trip_2022_09
sep <- sep %>%
  mutate(ride_length, ride_length = as.numeric(ride_length))%>%
  na.omit(ride_length)

oct <- trip_2022_10
oct <- oct %>%
  mutate(ride_length, ride_length = as.numeric(ride_length))%>%
  na.omit(ride_length)

nov <- trip_2022_11
nov <- nov %>%
  mutate(ride_length, ride_length = as.numeric(ride_length))%>%
  na.omit(ride_length)

dec <- trip_2022_12
dec <- dec %>%
  mutate(ride_length, ride_length = as.numeric(ride_length))%>%
  na.omit(ride_length)
head(dec)
```

The monthly data frames were combined together to create a data frame observing 
the entire year of 2022. This data frame will be used for most of the analysis.

```{r yearly data, include = FALSE}
library(lubridate)
year = bind_rows(jan, feb, mar, apr, may, jun, jul, aug, sep, oct, nov, dec)
```

New columns were added to the year data set including the year, month, and date
for each bike ride. 

```{r yearly data analysis, include = FALSE}
year$date <- as.Date.character(year$started_at, format = "%m/%d/%Y")
year$month <- format(as.Date(year$date), "%m")
year$day <- format(as.Date(year$date), "%d")
year$yr <- format(as.Date(year$date), "%Y")
```


## Analysis


### Monthly Data


The following graph shows that there are less riders in the Winter months compared 
to Spring and Summer. This observation can be seen for both casual and annual 
members.


```{r Analyze number of bike rides, echo = FALSE}
ggplot(data = year) + 
  geom_bar(aes(x = month, fill = member_casual), position = 'dodge') +
  labs(title = "Bike Ride Frequency Each Month in 2022", x = "Month",
       y = "Number of Rides", fill = "Member Type", 
       caption = "Data source: Motivate International Inc")
#Motivate International Inc
```


### Daily Data
The following graph shows that casual members rode more on the weekends
compared to the weekdays. Annual members appeared to rode more on the weekdays
compared to the the weekends. However, overall we see that annual riders
bike more consistently across the whole week compared to casual riders.


```{r day of week bike rides, echo = FALSE}
ggplot(data = year) + 
  geom_bar(aes(x = day_of_week, fill = member_casual), position = 'dodge') +
  labs(title = "Bike Ride Frequency Each Day in 2022", x = "Day of Week",
       y = "Number of Rides", fill = "Member Type", 
       caption = "Data source: Motivate International Inc",
       subtitle = "1 = Sunday; 7 = Saturday") +
       scale_x_continuous(labels = 1:7, breaks = 1:7)
```


### Average Ride Length: Casual VS Member


Compute the average ride length for both casual and annual members. The plots
below show the average ride length for the year, month, and day of the week.
The pattern we find is that the average ride length across the whole year,
month, and days of the week. The average ride length of the year for casual
riders was 23.84 minutes and annual riders was 12.45 minutes.


```{r summary data, echo = FALSE}
summary_data <- year %>%
  group_by(member_casual)%>%
  summarise(avg_ride_time = mean(ride_length)/60)

monthly_avg <- year %>%
  group_by(month, member_casual)%>%
  summarise(avg_ride_time = mean(ride_length)/60 , .groups = 'drop')

daily_avg <- year %>%
  group_by(day_of_week, member_casual)%>%
  summarise(avg_ride_time = mean(ride_length)/60 , .groups = 'drop')

ggplot(summary_data) +
  geom_col(aes(x=member_casual, y=avg_ride_time, fill = member_casual), show.legend = F) +
  labs(title = "Average Ride Time in 2022", y = "Ride Length (in minutes)", x = "Member Type",
       caption = "Data sourcs:Motivate International Inc")

ggplot(monthly_avg) +
  geom_col(aes(x=month, y=avg_ride_time, fill = member_casual),position = 'dodge') +
  labs(title = "Average Ride Time Each Month in 2022", y = "Ride Length (in minutes)", x = "Month",
       fill = "Member Type",
       caption = "Data source:Motivate International Inc")

ggplot(daily_avg) +
  geom_col(aes(x=day_of_week, y=avg_ride_time, fill = member_casual),position = 'dodge') +
  labs(title = "Average Ride Time Each Day in 2022", y = "Ride Length (in minutes)", x = "Day Of Week",
       fill = "Member Type",
       caption = "Data Source:Motivate International Inc",
       subtitle = "1 = Sunday;7 = Saturday") +
  scale_x_continuous(labels = 1:7, breaks = 1:7)

year %>%
  group_by(member_casual)%>%
  summarise(avg_ride_time = mean(ride_length)/60)
```


### Electric Vs Classic Bikes


We investigate the types of bikes ridden by casual and annual members across the
year and each day. We find that annual riders significantly use classic bikes 
more than electric bikes while casual riders used around the same amount of
electric and classic bikes the entire year. However, we find that the casual
riders use classic and electric bikes more on the weekends. This reflects the 
general behavior of casual riders riding bikes more on the weekends.


```{r electric and classical bikes, echo = FALSE}
year %>%
  filter(rideable_type!="docked_bike")%>%
  ggplot() +
  geom_bar(aes(x = month, fill = member_casual), position = 'dodge') + 
  facet_wrap(~rideable_type) + labs(title = "Types of Bikes Used Each Month",
                                    subtitle = "Classic and Electric Bikes",
                                    caption = "Data source:Motivate International Inc",
                                    fill = "Member Type",
                                    x = "Month")+
  scale_fill_brewer(palette="Dark2")

year %>%
  filter(rideable_type!="docked_bike")%>%
  ggplot() +
  geom_bar(aes(x = day_of_week, fill = member_casual), position = 'dodge') + 
  facet_wrap(~rideable_type) + 
  labs(title = "Types of Bikes Used Each Day",
      subtitle = "Classic and Electric Bikes",
      caption = "Data source:Motivate International Inc",
      x = "Day of Week",
      fill = "Member Type")+
      scale_fill_brewer(palette="Dark2")
```




### Popular Bike Routes


Investigated popular routes by both casual and annual members. This takes account
of bike rides where start and end stations are the same to consider all 
routes.

Below we find some of the most popular routes. We observe that casual member 
routes are mostly starting in tourist spots such as Millennium Park. Meanwhile,
we find that casual members have more routes starting outside of tourist 
hot spots and are more spread out around the Chicago area.


```{r popular routes, echo= FALSE}
routes <- year %>%
  group_by(start_station_name, end_station_name,member_casual)%>%
  summarise(total = n(), avg_time = mean(ride_length)/60, .groups = 'drop') %>%
  arrange(-total)

casual_route <- routes %>%
  filter(member_casual == "casual")
head(casual_route)

member_route <- routes %>%
  filter(member_casual == "member")
head(member_route)
```


Created a csv file to import into Tableau and create a data visualization of 
Chicago and popular start stations and routes used by casual and annual
members. We can find that more casual riders are found in tourist areas such
as downtown or near famous landmarks. 


```{r csv file, include = FALSE}
stations <- year %>% 
filter(start_lng != end_lng & start_lat != end_lat) %>%
group_by(start_station_name,end_station_name,start_lng, start_lat, end_lng, end_lat, member_casual) %>%
summarise(total = n(),.groups="drop") %>%
filter(total > 250)
```


### Tableau Visualization

You can click here to observe the dashboard: [Chicago Bike Routes and Stations](https://public.tableau.com/app/profile/ryan6592/viz/ChicagoCyclitic/Dashboard1#1)



## Conclusion


Based on the analysis on the data of bike rides in 2022, we found the following:

* There was a higher number of casual riders than members on the weeekend
* On average, casual riders had a higher ride duration than members across the year
* Classic bikes and electric bikes are used in similar amount for both member types
* The popular casual rider stations/routes begin in tourist or recreational spots

We can draw the conclusion that although annual members ride more each day we,
find that casual riders use the bikes for longer periods, specifically, weekends.
In addition, the Spring and Summer seasons are where most bike rides are 
found.Furthermore, from observations of popular bike routes, casual riders are attracted
more to tourist locations like Millennium park, etc. Also, there was no clear
indication of what kinds of bikes casual members preferred (electric or classic).


We can assume that there might be a larger frequency of member bike rides because 
it can be used more for commuting to work from Monday through Friday. We also
take account the each ride ID is different but we do not know if it is the same
individual riding different times each day. However, we only wanted to observe 
the frequency of bike rides according to type of membership. 



In conclusion, some recommendations may include the following:

* Offer first 6 months for free for annual membership. This will be in handy during
the Spring and Summer months where there are more bike rides

* Digital marketing on travel websites or in partnership with travel agencies for
tourists coming to Chicago

* Create more advertising near popular start stations for casual riders; use
tourist incentives such as perks with annual membership

* Market towards residents or workers who commute to work; incentivize biking
for casual riders who go to work







